---
title: "Mid  term Exam"
author: "Amie Marie Flora DUSHIMUMUKIZA"
date: "2025-10-12"
output:
  pdf_document: default
  html_document: default
lecturer: ' Dr.Pacifique Nizeyimana'
module: R for Data Science
---
 **Project I: World Population Trend and CO2 Emission**
 
In this dataset project, we are interested in analyzing the relationship between population growth and CO2 emissions. We have data for world population and CO2
emissions. The population dataset contains different features such as county names, population density, population growth rate, population ranking, and world population
percentage.

**Tasks to perform:**

**4.1 Downloading data and load them to R studio**
```{r}
# Load World Population Dataset
world_population<-read.csv("C:/Users/HP/OneDrive/Desktop/Data_midterm_exam/world_population.csv")
head(world_population,10) 
```

```{r}
# Loading CO2 Emissions
CO2_emissions<-read.csv("C:/Users/HP/OneDrive/Desktop/Data_midterm_exam/co2_emission.csv")
head(CO2_emissions,10)
```
**4.2 Exploratory data analysis**

 **a)variable names of World Population Dataset**
 
```{r}
variable.names(world_population)
```
 **b)Displaying the top 5 rows**
 
```{r}
head(world_population,5)
```
 **c)Displaying the bottom 10 rows**
 
```{r}
tail(world_population,10)
```

 **d)The data type**

```{r}
str(world_population)
```
 **e)The shape of the dataset**
 
```{r}
nrow(world_population)  # Number of rows (observations)
ncol(world_population)  # Number of columns (variables)
```
 **f)Checking duplicate rows**

```{r}
sum(duplicated(world_population)) # The output shows that no duplicates in dataset of world Population
```
 **g)find the number of missing values in each column**
 
```{r}
colSums(is.na(world_population)) # The output shows that
```
 **h)Using a box plot to check if there are any outliers in the quantitative variables**
```{r}
# Installing and Running library
library(ggplot2)
library(reshape2)

# Selecting only numeric columns
numeric_data <- world_population[sapply(world_population, is.numeric)]

# Normalize numeric columns (0–1 scaling)
normalized <- as.data.frame(lapply(numeric_data, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

# Convert to long format for ggplot (Reshaping)
reshaped_data <- melt(normalized, variable.name = "Variable", value.name = "Value")

# Create the boxplot
ggplot(reshaped_data, aes(x = Variable, y = Value)) +
  geom_boxplot(fill = "green", color = "blue") +
  theme_minimal() +
  labs(
    title = "Boxplot of World Population Variables",
    x = "Variables",
    y = "Normalized Values (0–1)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
 **i)Dealing with outliers**

```{r}
# Function to remove outliers using IQR rule
remove_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  x[x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR)] <- NA
  return(x)
}
#numeric_data <- world_population[sapply(world_population, is.numeric)]

# Remove outliers in each numeric column
numeric_withno_outliers <- as.data.frame(lapply(numeric_data, remove_outliers))

# Normalize again after removing outliers
library(reshape2)
library(ggplot2)

numeric_withno_outliers_normalize <- as.data.frame(lapply(numeric_withno_outliers, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))

# reshaping
reshaped_data_withno_outliers <- melt(numeric_withno_outliers_normalize, variable.name = "Variable", value.name = "Value")

# Ploting boxplot
ggplot(reshaped_data_withno_outliers, aes(x = Variable, y = Value)) +
  geom_boxplot(fill = "lightgray", color = "darkgray") +
  theme_minimal() +
  labs(
    title = "Boxplot of World Population Variables (Outliers Removed)",
    x = "Variables",
    y = "Normalized Values (0–1)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

 **4.3 Generating new Variable by using World Population Dataset**
 
```{r}
# Running library
library(dplyr)

#Creating new valiable
extraced_variable<-select(world_population,Country.Territory,X2022.Population,Growth.Rate)
head(extraced_variable)
```
**Creating new variable of projected population in 2030**

```{r}
exponential_growth <- function(P, r, t) {
  X2030.Population <- P * exp(r * t)
  return(X2030.Population)
}

population_data <- extraced_variable %>%
   mutate(X2030.Population = X2022.Population * exp((Growth.Rate/100) * 8))
  
head(population_data)
```

 **4.4 Value extraction and plot**
 
**a)Based on 2022 population, extracting top ten countries with high population number**

```{r}
library(dplyr)
# Extracting top 10 countries
top10_countries <- extraced_variable %>%
  select(-Growth.Rate) %>%        # remove the column of Growth.Rate
  arrange(desc(X2022.Population))  # sort by population
head(top10_countries, 10)
```
**b)using ggplot2 to perform the following task:**

1. Use an appropriate graph to present top 10 most populous counties and their
 population number during 2022.
```{r}
# Running library
library(ggplot2)

# Take top 10 and arrange them in descending order of population
top10 <- top10_countries %>%
  arrange(desc(X2022.Population)) %>%
  head(10)

# Create the plot
ggplot(top10, aes(x = reorder(Country.Territory, X2022.Population),
                  y = X2022.Population)) +
  geom_bar(stat = "identity", fill = "pink") +
  coord_flip() +
  labs(title = "Top 10 Most Populous Countries (2022)",
       x = "Country",
       y = "2022 Population") +
  theme_minimal(base_size = 14)
```
 2. Show the trend in their population number since 1990-2022 by using appropriate graph.

```{r}
# Selecting variables from 1990 to 2022 population number
world_population1990_2022<-world_population[,c("Country.Territory","X1990.Population","X2000.Population","X2010.Population","X2015.Population","X2020.Population","X2022.Population")]
world_population1990_2022
```

```{r}
# Selecting top 10 countries
top10_countr1990_2022 <- world_population1990_2022[world_population1990_2022$Country %in% c("Mexico", "Russia", "Bangladesh", "Brazil", "Nigeria","Pakistan","China","India","United States","Indonesia" ), ]
top10_countr1990_2022
```
**Ploting trend of the 10 ten population number since 1990-2022**

```{r}
# Load library

library(tidyverse)
data <- top10_countr1990_2022

# Convert from wide to long format
data_long <- data %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "Population"
  ) %>%
  mutate(Year = as.numeric(gsub("X|\\.Population", "", Year)),
         Population_Millions = Population / 1e6)  # convert to millions

# Plot population trend
ggplot(data_long, aes(x = Year, y = Population_Millions, color = Country.Territory)) +
  geom_line(size = 1.3) +
  geom_point(size = 2) +
  labs(
    title = "Population number Trend (1990–2022)",
    x = "Year",
    y = "Population (Millions)",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, linewidth = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")

```
3.Use CO2 Emissions Around the World dataset to extract emission of 10 most
 populous countries. Use an appropriate graph to show their emission trend
 since 1990-2019.
 
```{r}
# Loading CO2 Emissions
CO2_emissions<-read.csv("C:/Users/HP/OneDrive/Desktop/Data_midterm_exam/co2_emission.csv")
```
 
Extract 10 most populous countries in emission of CO2
```{r}
top10_CO2_emission <- CO2_emissions[CO2_emissions$Country %in% c("Mexico", "Russia", "Bangladesh", "Brazil", "Nigeria","Pakistan","China","India","United States","Indonesia", "Russian Federation" ), ]
top10_CO2_emission
```

**Extracting emission variables from X1990 t0 X2019 of the top 10 most populous countries**
```{r}
Top10_Co2_emission_1990_2019<-top10_CO2_emission[c(1,5:34)]
Top10_Co2_emission_1990_2019
```
```{r}
# Load required libraries
library(tidyverse)

data1 <- Top10_Co2_emission_1990_2019  

# Reshape data from wide to long
data_long <- data1 %>%
  tidyr::pivot_longer(
    cols = starts_with("X"),
    names_to = "Year",
    values_to = "CO2_Emissions"
  ) %>%
  mutate(Year = as.numeric(sub("X", "", Year)))

# Plot
ggplot(data_long, aes(x = Year, y = CO2_Emissions, color = Country.Name)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "CO2 Emissions trend by top 10 populous Countries ",
    x = "Year",
    y = "CO2 Emissions",
    color = "Country"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )
```
 **4.5 Correlation Analysis**
 
 By Using the population dataset, find correlations between Area , Density ,
 Growth rate, and World Population Percentage by using both numerical values
 and heatmap.
 
```{r}
# Load required library

library(corrplot)
# Creating correlation matrix of the specified variables

cor_matrix <- cor(world_population[, c("Area..km..", "Density..per.km..",
                                       "Growth.Rate", "World.Population.Percentage")])

# Plotting correlation heatmap

corrplot(cor_matrix, method = "color", addCoef.col = "black")

title(main = "Correlation Heatmap of some World Population Variables", 
      line = 2,  # adjust vertical position
      cex.main = 1.2,  # adjust title size
      col.main = "darkblue")  # title color
```

**2. Merge the World Population Dataset and CO2 Emissions around the World on country name** 
 
```{r}
#Merge datasets on country name
merged_data <- world_population %>%
  inner_join(CO2_emissions, by = c("Country.Territory" = "Country.Name")) 
merged_data
```
**Extracting 2022 population and 2019 CO2 emissions.**

```{r}
extract <- merged_data[,c("Country.Territory","X2022.Population","X2019")]
extract
```
 
**checking the missing values and Cleaning the extracted dataset** 
```{r}
colSums(is.na(extract))
```
**Removing missing values**

```{r}
extract_cleaned<-na.omit(extract) # I decided to remove missing value because the number of missing values is small in proportion.
extract_cleaned
```

**Visualizing relationships between population number and CO2 emissions**

```{r}
ggplot(extract_cleaned, aes(x = X2022.Population, y = X2019)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Relationship between Population (2022) and CO2 Emissions (2019)",
       x = "Population in 2022",
       y = "CO2 Emissions in 2019 (metric tons per capita)") +
  theme_minimal()
```
**Here i decided to use Log base of 10 to transform compressed large values and spreads out smaller ones, making the data more normally distributed and easier to interpret and visualize relationship**.

```{r}
library(ggplot2)
ggplot(extract_cleaned, aes(x = log10(X2022.Population), y = log10(X2019))) +
  geom_point(color = "darkblue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dotted") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Logs Relationship Between Population and CO2 Emissions",
    x = "Log10(Population in 2022)",
    y = "Log10(CO2 Emissions in 2019)")
```

**Performing Statistical Analysis**

```{r}
correlation <- cor(extract_cleaned$X2022.Population, extract_cleaned$X2019, method = "pearson")
print(paste("Correlation between Population and CO2 Emissions is:", round(correlation, 2)))

```
**Descriptive Statistics of my data (Summary)**
```{r}
summary(extract_cleaned)
```
**Linear regression model** 
```{r}
model <- lm(X2019~ X2022.Population, data = extract_cleaned)
summary(model)

```

**4.6 Comparing CO2 Emissions in continents**

```{r}
# Load required packages
library(dplyr)
library(ggplot2)

# The dataset to use
data2 <- merged_data
  
# Cleaning and rename relevant columns
data <- data2 %>%
  rename(
    Country = Country.Territory,
    Continent = Continent,
    Population_2022 = X2022.Population,
    CO2_2019 = X2019
  )

# Removing missing values if any
data <- na.omit(data)

# Summarize total CO2 emissions by continent
continent_summary <- data %>%
  group_by(Continent) %>%
  summarise(Total_CO2_2019 = sum(CO2_2019, na.rm = TRUE))
ggplot(continent_summary, aes(x = reorder(Continent, -Total_CO2_2019),
                              y = Total_CO2_2019,
                              fill = Continent)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Total_CO2_2019, 2)),   # show total CO₂
            vjust = -0.5, size = 4) +
  labs(
    title = "Total CO2 Emissions by Continent (2019)",
    x = "Continent",
    y = "Total CO2 Emissions (metric tons per capita)",
    fill = "Continent"
  ) +
  theme_minimal()

geom_text(aes(label = scales::comma(round(Total_CO2_2019, 1))),
          vjust = -0.5, size = 4)
```
**Based on the graph generated, answer the following:**

 1. What is the first continent with high 2019 CO2 emission?
 
 - The first continent with high 2019 CO2 emission is: **Europe**
 
 2. What is the third continent in terms of emitting CO2? What is their total
 emission(metric tons per capita)?
 
 - The third continent in terms of emitting CO2 is: **North America**
 - Total emission is **79.57 metric tons per capita**
 
 3. What is the last continent in terms of emitting CO2? What their total emis
sion(metric tons per capita)?

- The last continent in terms of emitting CO2 is: **South America**
- Total emission is **29.42 metric tons per capita**



